---
title: "STAT 423 Final Projct"
author: "Andrew Sousa"
date: "2025-02-19"
output: pdf_document
---

```{r}
library(tidyverse)
library(MASS)
library(dplyr)
```

Model 1a) Simple Linear Regression
For our first model, we are going to use a simple linear regression to try and predict divorce rates in Washington. To do this, we will look at two variables:

- Predictor 1 : GDP: The change in percent in gross domestic product in the state of Washington in any given year 
- Estimated Value: Divorce rate: the rate of divorce per 1000 couples in the state of Washington in any given year.

Using this simple model, we would have an linear regression model of:

$$ \widehat{\text{Divorce Rate}_i} = \widehat{\beta}_0 + \widehat{\beta}_1\text{gdp}_i + \epsilon_i$$

Data cleanup, combining gdp and divorce rate into one dataframe
```{r}
gdp <- read.csv("WA_GDP.csv")
divorce <- read.csv("divorce_rates.csv")

divorce <- divorce %>% filter(X == "Washington")

gdp <- gdp[1,]

gdp <- dplyr::select(gdp, c(-1, -3, -4, -5, -6, -7, -8))
gdp <- gdp %>% rename(stat = GeoName)
gdp$stat <- "gdp"
gdp <- dplyr::select(gdp, c(-2,-3,-27))

divorce <- dplyr::select(divorce, c(-27,-28,-29,-30,-31,-32))
divorce <-dplyr::select(divorce, c(X, X2000, X2001, X2002, X2003, X2004, X2005, X2006, X2007, X2008, X2009, X2010, X2011, X2012, X2013, X2014, X2015, X2016, X2017, X2018, X2019, X2020, X2021, X2022))
divorce <- divorce %>% rename(stat = X)
divorce$stat <- "divorce_rate"

gd_data <- rbind(gdp, divorce)
gd_data <- gd_data %>% pivot_longer(cols = c(X2000, X2001, X2002, X2003, X2004, X2005, X2006, X2007, X2008, X2009, X2010, X2011, X2012, X2013, X2014, X2015, X2016, X2017, X2018, X2019, X2020, X2021, X2022),
                                    names_to = "year",
                                    values_to = "value") %>%
  pivot_wider(names_from = stat, values_from = value)

gd_data <- gd_data %>% mutate(gdp = as.numeric(gdp),
                              divorce_rate = as.numeric(divorce_rate))

head(gd_data)
```


To start, we will create a basic model, with no transformations to see the initial relationship
```{r}
gd_model <- lm(data = gd_data, formula = divorce_rate ~ gdp)
gd_resid <- resid(gd_model)

ggplot(data = gd_data) +
  geom_point(mapping = aes(x = gdp, y = divorce_rate)) +
  geom_smooth(mapping = aes(x = gdp, y = divorce_rate), method = "lm") +
  labs(title = "Scatterplot of Change in GDP vs Divorce Rate in Washington",
       x = "Change in GDP (%)",
       y = "Divorce Rate (per 1000")


ggplot() +
  geom_histogram(mapping = aes(x = gd_resid), bins = 10) +
  labs(title = "Historgram of Residuals in SLR Model",
       x = "SLR Residuals")

summary(gd_model)
plot(gd_model)
```
An initial model looks to have many problems. GDP and Divorce Rate don't seem to follow any linear pattern, with many points falling well outside our best-fit line. Additionally, our residual plot shows a patterb in residuals, increasing until a fitted value of 3.85 before falling down again. A QQ-plot shows nearly half our residuals falling off the Normal line, meaning our residuals are not normally distributed. 

Alongside a p-value of 0.106 for gdp, and an $R^2$ of 0.1199, without transformations, GDP does a poor job of predicting Divorce Rate in Washington. We will try some transformations to see if the model can be improved, or conclude that we need more variables.

Transformation 1) rate reduction + log
To try and decrease the spread of our divorce rate to better fit GDP change, we will turn divorce rate into a percent like GDP, and take the log of it. + 1 will be added to each to avoid negative logs 
```{r}
gd_log <- data.frame("gdp" = gd_data$gdp,
                     "divorce_rate" = log(gd_data$divorce_rate/1000 + 1))
gd_log_model <- lm(data = gd_log, formula = divorce_rate ~ gdp)
gd_log_resid <- resid(gd_log_model)

ggplot(data = gd_log) +
  geom_point(mapping = aes(x = gdp, y = divorce_rate)) +
  geom_smooth(mapping = aes(x = gdp, y = divorce_rate), method = "lm") +
  labs(title = "Scatterplot of Change in GDP vs log(Divorce Rate/1000 in Washington)",
       x = "Change in GDP (%)",
       y = "log(divorce rate % + 1)")

ggplot() +
  geom_histogram(mapping = aes(x = gd_log_resid), bins = 10) +
  labs(title = "Historgram of Residuals in SLR Model",
       x = "SLR Residuals")

plot(gd_log_model)
summary(gd_log_model)
```
A log transformation seems to do little to change the relationship between GDP and divorce rate, with the same violations in our residuals and similar layout of scatterplots.     

Transformation 2) square roots
To try and lower the spread, we'll now take the square root of each rate. Like last time, we'll turn divorce rate into an actual percentage, and add 1 to each to avoid negative square roors 
```{r}
gd_sqrt <- data.frame("gdp" = gd_data$gdp,
                     "divorce_rate" = sqrt(gd_data$divorce_rate/1000 + 1))
gd_sqrt_model <- lm(data = gd_sqrt, formula = divorce_rate ~ gdp)
gd_sqrt_resid <- resid(gd_sqrt_model)

ggplot(data = gd_sqrt) +
  geom_point(mapping = aes(x = gdp, y = divorce_rate)) +
  geom_smooth(mapping = aes(x = gdp, y = divorce_rate), method = "lm") +
  labs(title = "Scatterplot of Change in GDP vs sqrt(Divorce Rate/1000 in Washington)",
       x = "Change in GDP (%)",
       y = "sqrt(divorce rate % + 1)")

ggplot() +
  geom_histogram(mapping = aes(x = gd_log_resid), bins = 10) +
  labs(title = "Historgram of Residuals in SLR Model",
       x = "SLR Residuals")

plot(gd_sqrt_model)
summary(gd_sqrt_model)
```
Again, a square root does little to change the relationship or any distributions of our residuals.

Transformation 3) Box-Cox
```{r}
boxcox(gd_model, lambda = seq(-2, 2, 0.1))
#closish to 0, but to try some transformation, we'll use the exact value from the boxcox likelihood

gd_t <- data.frame(gdp = (gd_data$gdp/100 + 1)^-0.15,
                    divorce_rate = (gd_data$divorce_rate/1000 + 1)^-0.15)

gd_data
gd_t_model <- lm(data = gd_t, formula = divorce_rate ~ gdp)
summary(gd_t_model)

ggplot(data = gd_t) +
  geom_point(mapping = aes(x = gdp, y = divorce_rate)) +
  geom_smooth(mapping = aes(x = gdp, y = divorce_rate), method = "lm")

plot(gd_t_model)
```
Not even the box-cox method improve our relationship, with the different but still problematic scatterplot and residual violations. It seems that transformations can't fix the relationship, but for one last look, we will do an anova test, comparing the models to a null model at significance level $\alpha$ = 0.05.

Anova test
```{r}
anova(gd_model)
anova(gd_log_model)
anova(gd_sqrt_model)
anova(gd_t_model)
```
For each model, we have p-values above 0.1, well above our desired significance level of 0.05. With this, we don't have statistically significant evidence that our SLR models are any better at predicting divorce rate than just a null. As such, it seems definitive that GDP alone is not enough to accurately predict divorce rate, and will need to look at more variables in future models.  


