---
title: "STAT 423 Final Report"
author: "Andrew Sousa, Shelby Ames, Malia Schomaker, Sebastian Villarroel"
date: "2025-03-08"
output: pdf_document
---

``{r label = "libraries", eval = TRUE, echo = FALSE}
library(tidyverse)
library(MASS)
library(dplyr)
library(conflicted)
conflicts_prefer(dplyr::filter)
conflicts_prefer(dplyr::select)
library(gridExtra)
```
# Introduction

# Questions
With this study, we look to answer four questions:

## 1) What are the relationships between GDP and divorce rates?
## 2) What is the best economic predictor for divorce rates?  
## 3) Would a population’s relationship with their spouse be a good indicator for the economy?
## 4) How do our relationships change when personal income, GDP, and expenditures change?

# Methods of Analysis
```{r label = "code", eval = TRUE, echo = FALSE}

```


# Models
```{r label = "data", eval = TRUE, echo = FALSE}
gdp <- read.csv("WA_GDP.csv")
divorce <- read.csv("divorce_rates.csv")
```

To answer our questions, we have created a multitude of models to find the relationships between our variables

## Model 1b) Simple Linear Regression Model (GDP ~ Divorce Rate)
For a baseline assessment, we will now look at a simple model, analyzing the relationship between GDP and divorce rate. Simply,

$$\text{Divorce Rate} = \beta_o + GDP + \epsilon$$

where GDP = percent change in total GDP from prior year and Divorce Rate = number of divorces per 1000.


# Results

## Model 1b) Struggles with simple linear regression
```{r label = "slr_data", eval = TRUE, echo = FALSE}
divorce <- divorce %>% filter(X == "Washington")

gdp <- gdp[1,]

gdp <- dplyr::select(gdp, c(-1, -3, -4, -5, -6, -7, -8))
gdp <- gdp %>% rename(stat = GeoName)
gdp$stat <- "gdp"
gdp <- dplyr::select(gdp, c(-2,-3,-27))

divorce <- dplyr::select(divorce, c(-27,-28,-29,-30,-31,-32))
divorce <- dplyr::select(divorce, c(X, X2000, X2001, X2002, X2003, X2004, X2005, X2006, X2007, X2008, X2009, X2010, X2011, X2012, X2013, X2014, X2015, X2016, X2017, X2018, X2019, X2020, X2021, X2022))
divorce <- divorce %>% rename(stat = X)
divorce$stat <- "divorce_rate"

gd_data <- rbind(gdp, divorce)
gd_data <- gd_data %>% pivot_longer(cols = c(X2000, X2001, X2002, X2003, X2004, X2005, X2006, X2007, X2008, X2009, X2010, X2011, X2012, X2013, X2014, X2015, X2016, X2017, X2018, X2019, X2020, X2021, X2022),
                                    names_to = "year",
                                    values_to = "value") %>%
  pivot_wider(names_from = stat, values_from = value)

gd_data <- gd_data %>% mutate(gdp = as.numeric(gdp),
                              divorce_rate = as.numeric(divorce_rate))

gd_data <- gd_data %>% select(gdp, divorce_rate)
```

Despite many transformations and tests, a simple linear regression struggles massively to accurately predict divorce rate in Washington state.

To start, we will create a basic model, with no transformations to see the initial relationship
```{r, label = "basic_slr_model", eval = TRUE, echo = FALSE, fig.width = 4, fig.height = 4, fig.align='center'}
gd_model <- lm(data = gd_data, formula = divorce_rate ~ gdp)
gd_resid <- resid(gd_model)

plot(gd_data, main = "Scatterplot of GDP vs Divorce Rate", xlab = "GDP", ylab = "Divorce Rate")
abline(gd_model)

hist(x = gd_resid, main = "Histogram of Divorce Rate ~ GDP Residuals")
```
An initial model looks to have many problems. GDP and Divorce Rate don't seem to follow any linear pattern, with many points falling well outside our best-fit line. Additionally, our residual plot shows a pattern in residuals, increasing until a fitted value of 3.85 before falling down again. A QQ-plot shows nearly half our residuals falling off the Normal line, meaning our residuals are not normally distributed. 

Alongside a p-value of 0.106 for gdp, and an $R^2$ of 0.1199, without transformations, GDP does a poor job of predicting Divorce Rate in Washington. We will try some transformations to see if the model can be improved, or conclude that we need more variables.

### Transformations
We are trying three transformations: log, square root, and lastly using the box-cox method
```{r label = "slr_trans_models", eval = TRUE, echo = FALSE, fig.width = 4, fig.height = 4}
# log transformation
gd_log <- data.frame("gdp" = gd_data$gdp,
                     "divorce_rate" = log(gd_data$divorce_rate+ 0.001))

# square root transformation
gd_sqrt <- data.frame("gdp" = gd_data$gdp, 
                      "divorce_rate" = sqrt(gd_data$divorce_rate))

# box-cox derived transformation
gd_boxcox <- data.frame("gdp" = gd_data$gdp,
                        "divorce_rate" = (gd_data$divorce_rate)^(-0.15))

gd_log_model <- lm(data = gd_log, formula = divorce_rate ~ gdp)
gd_sqrt_model <- lm(data = gd_sqrt, formula = divorce_rate ~ gdp)
gd_boxcox_model <- lm(data = gd_boxcox, formula = divorce_rate ~ gdp)
```

We will do an anova test, to see if any of our models is any better than our original in predicting divorce rate

```{r label = "slr_anova", eval = TRUE, echo = FALSE}
anova(gd_log_model)
anova(gd_sqrt_model)
anova(gd_boxcox_model)
```
For each model, we have p-values above 0.1, well above a desired significance level of 0.05. With this, we don't have statistically significant evidence that our SLR models are any better at predicting divorce rate than just a null. As such, it seems definitive that GDP alone is not enough to accurately predict divorce rate, and will need to look at more variables in future models. 

# Conclusions

# References

# Appendix
``` {r ref.label = "code", eval = FALSE}

```
# Loading in packages
```{r ref.label = "libraries", eval = FALSE, echo = TRUE}

```

# Reading the data into the file
```{r ref.label = "data", eval = FALSE, echo = TRUE}
```

# Turning GDP and Divorce Rate into a single file for SLR
```{r ref.label = "slr_data", eval = FALSE, echo = TRUE}

```

# Creating basic model for SLR
```{r ref.label = "basic_slr_model", eval = FALSE, echo = TRUE}

```

# Transformations to SLR Model
```{r ref.label = "slr_trans_models", eval = FALSE, echo = TRUE}

```

# SLR Model Anovas
```{r ref.label = "slr_anova", eval = FALSE, echo = TRUE}

```
