---
title: "STAT 423 Final Report"
author: "Andrew Sousa, Shelby Ames, Malia Schomaker, Sebastian Villarroel"
date: "2025-03-08"
output: pdf_document
---

# Introduction

# Questions
With this study, we look to answer four questions:

## 1) What are the relationships between GDP and divorce rates?
## 2) What is the best economic predictor for divorce rates?  
## 3) Would a population’s relationship with their spouse be a good indicator for the economy?
## 4) How do our relationships change when personal income, GDP, and expenditures change?

# Methods of Analysis
```{r label = "code", eval = TRUE, echo = FALSE}

```


# Models
To answer our questions, we have created a multitude of models to find the relationships between our variables

## Model 1a) Simple Linear Regression Model (GDP ~ Divorce Rate)
For a baseline assessment, we will first look at a simple model, analyzing the relationship between GDP and divorce rate. Simply,

$$\text{Divorce Rate} = \beta_o + GDP + \epsilon$$


# Results

# Conclusions

# References

# Appendix
``` {r ref.label = "code", eval = FALSE}

```

